Efficient Data Provenance Analysis for Concurrent Programs

### 2013-11-15 paper.tex
committing a draft that includes the data from the collisions experiment.  back on 10 pages.  fingers crossed.

A value's provenance is a description of the reason that it is stored in a
variable. Provenance information helps answer the question: "Why does this
variable contain this value?", which is central to debugging. Debugging a
concurrent program requires understanding how threads interact, and since
communication happens via reads and writes to shared variables, {\em data
provenance information is essential}. Provenance can take on many forms, for
example, a value's provenance could be the instruction that wrote it, the
function that returned it, and in the context of concurrent programs, the
thread that produced it.

@@

Collecting debugging information, like data provenance, from production
systems is particularly important for concurrent software since bugs
manifest as failures rarely. When a production system fails and a programmer
receives a bug report, reproducing the reported failure to collect provenance
information may be difficult or impossible. It is essential to collect as much
information as performance margins permit from production, to reduce the need
for time-consuming failure reproduction.

@@

In this section we discuss debugging with data dependence information, using an
example, in order to motivate collecting last writer slices in deployed
systems.  We then provide context for \ctraps by discussing some more
sophisticated concurrency analyses that rely on the kind of dependence
information we are focusing on in this work.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug2.pdf}
\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
\end{figure}

Another important trait of last writer slices illustrated by
Figure~\ref{fig:jdklws} is that when the assertion fails at {\bf B}, stopping
the execution, the \lwt is saved and can be sent to the programmer with the
core dump.  Capturing good debugging information is especially important for
concurrency bugs that manifest very rarely or in the presence of only peculiar
real-world inputs.  A primary goal of this work is to show that we can collect
last writer slices with overheads that are low enough for production use so
they can be included pervasively in bug reports.

@@

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is written.
A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
compiler inserts \lwt updates before every write operation (for completeness)
and only before write operations (for soundness).  Formally, given a program,
$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
sound, complete, and thus correct.

Next we discuss two nuances of our design: (1) the interaction of our \lwt
implementation with code containing data-races; and (2) the correctness of the
\lwt when a program addresses data that overlap in memory.

@@

\paragraph{Non-overlapping Memory References} 
We assume each written memory location corresponds to a non-overlapping memory
region because our \lwt design is {\em granularity oblivious}.  We discuss this
issue informally due to space constraints.  Each \lwt entry corresponds to a
memory location referenced by a program instruction.  Two instructions might
reference different locations that overlap because the referenced locations
have different granularities ({\em e.g.}, accessing a byte that is part of a
previously accessed word).  When that happens, both accesses update different
\lwt entries, because they reference different locations.  However, the
overlapping region then has two different entries reflecting the last {\em two}
updates to its parts rather than just its last write.

The \lwt remains correct, assuming the program only accesses non-overlapping
regions.  In programs that access overlapping regions, there is always some
entry in the \lwt that correctly reflects each memory location's last writer.
However, the entry for a coarse-grained region may be obsolete because of an
access to an overlapping sub-region.

@@

\paragraph{Communication Graph Implementation}
Communication graphs are the basis of several prior debugging
techniques~\cite{recon, bugaboo, defuse}.  We implement communication graph
collection using \ctraps.  When a trap is delivered, our implementation records
communication graph edges composed of the code point of the last writer to the
location being accessed and the code point of the trapping access.  The set of
recorded edges is the tool's output.  Our \ctraps implementation of
communication graphs took about 50 lines of code (including output formatting
code).

@@

\section{System Implementation}

We implemented last writer slicing and \ctraps.  We built our compiler support
as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
escape analysis support to prune instrumentation on accesses to non-escaping memory
locations.  Our compiler instruments calls to {\tt free} and {\tt delete} as
writes to the pointer being deallocated.  Our analysis handles all code
compilable using this version of GCC except for a few cases: we do not handle
accesses that compile to GCC {\tt BITFIELD\_REFERENCE} IR types because these
are undocumented, we do not handle inlined assembly instructions, and we do not
handle C++ exceptions.  Note that these limitations of our research prototype
are not fundamental and the limited cases are uncommon.

We built our runtime system from scratch.  We implemented the LWT as a fixed
size 2GB array of 64-bit words, fixing our memory overheads.  Each entry packs a program point and a thread
identifier into the 64-bit word.  By default, program points are single
instruction addresses, but we also support limited (2-address)
context-sensitivity by packing the current instruction's address and the
nearest return address into a 64-bit LWT entry. When an access to a memory
location occurs, we index into the LWT with the lower bits of the location's
address.  Our prototype implementation uses a lossy resolution policy for hash
collisions.  We use this policy in our prototype because it is faster and
simpler than a chaining concurrent hash table implementation.  Collisions are
extremely rare, so this simplification is unlikely to be a problem even in a
more industrial strength implementation.

\subsection{Release}
We released our implementation, free and open-sourced, on the
web~\cite{ctrapsrelease} and published it to the GCC plugin repository.
Since its release, we have seen contributions to our codebase from members of
the GCC development community and downloads from interested developers.

@@

The results in Table~\ref{tab:bugs} show that for both atomicity and ordering
violation bugs, last writer slices are effective.  In all cases, we found that
the last writer slice revealed the provenance of at least some data involved in
the failure, so last writer slicing was effective.  We found that none of the
failures we looked at were trivially debuggable.  This finding is consistent
with the difficulty of debugging concurrency bugs -- the core dump does not
reveal the interaction between threads that led to a failure, only the result
of the interaction, which leads to the failure.

Last writer slices were useful for helping understand how to fix the bug in
all but one case and useful at least for understanding the bug in the
remaining case, {\tt apache2}.  The main reason last writer slices are helpful
is that they reveal the connection between code running in different threads
that interacts to cause a failure.  Without last writer slices, programmers
have no easy way of connecting these otherwise disparate parts of a program.
The case studies in Section~\ref{sec:eval:debugging:cases} show in more detail
how last writer slices are useful for debugging.

%For the remaining case, {\tt apache2}, last writer slices would
%probably help with debugging, but it is less obvious because this bug is a
%complex, multi-variable bug.  We provide more detail on {\tt apache2} in
%Section~\ref{sec:eval:debugging:cases}.
Our evaluation shows that bad value origin tracking~\cite{badapples} is
definitely helpful in one case where last writer slices is useful and may help
in one other case.  For a use-after-delete bug in {\tt pbzip2}, bad value
origin tracking helps.  For this bug, when one thread deletes a variable, it
makes that variable unusable and bad value origin tracking would track that
value.  When the failure occurs, the programmer would then know why the value
was deleted.

@@

We used last writer slices to debug an atomicity violation in MySQL-4.0.12 that
is a critical security vulnerability.  Figure~\ref{fig:mysqllws} shows the
buggy code.  Thread 1 rotates the database's log file.  Log rotation should be
atomic, but the lack of synchronization allows Thread 2 to run its insert
during log rotation.  In that case, the insert query not logged, manifesting
the failure.  We added an {\tt abort} statement to the non-logging case because
this behavior is a security vulnerability that the bug report describes as
``critical''.  Adding the {\tt abort} is reasonable, as the bug report
describes the {\tt else} block as an error case.  Running in a debugger, a
programmer might equivalently use a breakpoint, rather than adding an
{\tt abort}.

@@

\paragraph{apache2}
We used last writer slices to debug an atomicity violation in Apache-2.2.9 and
we found that last writer slices help understand the bug, but not how to fix
it.  The bug involves a group of worker threads that handle web requests and a
network listener thread that passes requests to the workers.  The queue tracks
the number of idle workers.  When a worker becomes idle, it increments the
number of idlers and when the listener passes work to a worker, it decrements
that number.  The counter's increments and decrements are synchronized using
condition variables.  If there are no idlers, the listener waits for a signal.
When a worker becomes idle when no others are, it delivers a signal to the
listener.

The problem is that the listener does not check the condition governed by the
condition variable after being signaled on that condition variable.  As a
result, the listener can incorrectly decrement the idler count twice,
underflowing the unsigned value, which then causes an assertion to fail.  When
the failure occurs, the core dump shows the idler count underflowed, revealing
a primary symptom of the bug, but not providing much help.  We checked the last
writer slice for the idler count and it reported that the listener thread last
wrote the idler count at its decrement.  That information shows the programmer
{\em why} the underflow occurred, which is helpful for understanding the
failure.  However, we conclude that last writer slices are only somewhat 
useful in this case.  Solving this complex bug requires understanding not just
the connection between the underflow and the listener's decrement, but also the
synchronization protocol that coordinates that decrement and other accesses to
the variable.

@@

These high-level results support our main claim that the runtime overhead of
\ctraps is low enough for deployed systems for many applications, especially
for servers.  We now further analyze our observed overheads, illuminating
differences between the lower-overhead server programs and the higher-overhead
PARSEC programs.

@@

\subsubsection{Overheads Due to Conservative Escape Analysis}
\label{sec:eval:conservative}
{\tt swaptions} had the highest run time overhead in our tests.  We
investigated the cause and determined it is largely due to shortcomings in our
escape analysis.  We located the inner-loop function {\tt
HJM\_SimPath\_Forward\_Blocking}.  The function uses two local matrices that
are allocated through an external call.  Our escape analysis cannot determine
that the matrices are only used locally in that function, so it conservatively
assumes the matrices escape.  We manually inlined the matrix allocation calls
(and a matching deallocation calls), and escape analysis eliminated their
instrumentation.  The program's overhead dropped from about 420\% overhead for
\ctraps to about 300\% overhead; the change for last writer slicing was from
300\% to about 130\% overhead.  We also looked at an output matrix passed into
the inner-loop function.  The matrix is allocated and deallocated in the
inner-loop's caller, but is never shared.  Eliminating instrumentation on
accesses to this matrix yielded an overhead of about 46\% for \ctraps, and to
just 30\% for last writer slicing.

These manual improvements to the escape analysis reduced the run time overheads
for these workloads from being unacceptable for any production environment to
being reasonable for use in production systems.  We expect a more sophisticated
escape analysis would find these local accesses and achieve these performance
gains automatically.

\subsubsection{PARSEC vs. Servers}
\label{sec:eval:parsecserver}
In our tests, server programs were better able to tolerate the overhead of
last writer slices and \ctraps.   We discern three reasons for this difference.

@@

\subsection{Context-Sensitivity}
%{\bf TODO: Was this LWS or CTRAPS-NOOP or what?}
We briefly evaluated the cost of context-sensitivity using \ctraps with no-op
trap handlers.  With a bounded 2-address context, the average overhead for
server programs was 18\% and the average overhead for PARSEC was 145\%.  These
data show that a restricted form of context-sensitive analysis is possible in
\ctraps with overheads low enough for production.

@@

\subsubsection{Performance with \ctraps}

@@

\section{Conclusions}
Last writer slices are the first system to collect a general form of data provenance information for concurrent programs with overheads low enough for production use.

We developed system support that collects last writer slices with overheads
well within an acceptable range for production.  We showed that, during
debugging, the provenance information provided by last writer slices reveals
crucial connections between different threads' otherwise disparate regions of
code.  Understanding how code in different threads interacts is essential to
concurrency debugging and this work helps with that understanding.

>>>>

A value's provenance is a description of the reason that it is stored in a
variable. Provenance information helps answer the question: "Why does this
variable contain this value?", which is central to debugging. Debugging a
concurrent program requires understanding how threads interact, and because  
communication happens via reads and writes to shared variables, {\em data
provenance information is essential}. Provenance can take on many forms, for
example, a value's provenance could be the instruction that wrote it, the
function that returned it, and in the context of concurrent programs, the
thread that produced it.

@@

Efficiently collecting debugging information, like data provenance, from production
systems is particularly important for concurrent software since bugs
manifest as failures rarely. When a production system fails and a programmer
receives a bug report, reproducing the reported failure to collect provenance
information may be difficult or impossible. It is essential to collect as much
information as performance margins permit from production, to reduce the need
for time-consuming failure reproduction.

@@

In this section we discuss debugging with data dependence information, using an
example, in order to motivate collecting last writer slices in deployed
systems.  We then provide context for \ctraps by discussing some more
sophisticated concurrency analyses that rely on last writer information.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug2.pdf}
\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the buggy interaction between {\tt erase()} and {\tt append()}.}
\end{figure}

Another important trait of last writer slices illustrated by
Figure~\ref{fig:jdklws} is that when the assertion fails at {\bf B}, stopping
the execution, the \lwt is saved and can be sent to the programmer with the
core dump.  Capturing good debugging information is especially important for
concurrency bugs that manifest very rarely or in the presence of only peculiar
real-world inputs.  A primary goal of this work is to show that we can collect
last writer slices with very low overheads so they can be included pervasively
in bug reports.

@@

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For an
unwritten location $n$, $LWS[n] = (\emptyset,\emptyset)$.

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design updates $LWT[m]$ only when $m$ is written.
A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
compiler inserts \lwt updates before every write operation (for completeness)
and only before write operations (for soundness).  Formally, given a program,
$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
sound, complete, and thus correct.

Next we discuss two nuances of our design: (1) the interaction of our \lwt
implementation with data-races; and (2) the correctness of the
\lwt for accesses to overlapping memory regions.

@@

\paragraph{Non-overlapping Memory References} 
We assume written memory locations do not overlap in memory because our \lwt
design is {\em granularity oblivious}.  We discuss this issue informally due to
space constraints.  Each \lwt entry corresponds to a location that was written.
Two instructions might reference different locations that overlap because the
referenced locations have different granularities ({\em e.g.}, accessing a byte
that is part of a previously accessed word).  When that happens, both accesses
update different \lwt entries, because they reference different locations.
However, the overlapping region then has two different entries reflecting the
last {\em two} updates to its parts rather than just its last write.

The \lwt remains correct, assuming the program only accesses non-overlapping
regions.  In programs that access overlapping regions, there is always some
entry in the \lwt that correctly reflects each memory location's last writer.
However, the coarse region's entry may be stale because of an
access to an overlapping region.

@@

\paragraph{Communication Graph Implementation}
Communication graphs are the basis of several prior debugging
techniques~\cite{recon, bugaboo, defuse}.  We implement communication graph
collection using \ctraps.  When a trap is delivered, our implementation records
communication graph edges composed of the code point of the last writer to the
location being accessed and the code point of the trapping access.  The set of
recorded edges is the tool's output.  Our \ctraps implementation of
communication graphs took about 50 lines of code.

@@

\section{System Implementation}
\label{sec:implementation}
We implemented last writer slicing and \ctraps.  We built our compiler support
as a plugin for GCC ({\tt gcc-4.7}).  We used GCC's points-to and
escape analysis support to prune instrumentation on accesses to non-escaping memory
locations.  Our compiler instruments calls to {\tt free} and {\tt delete} as
writes to the pointer being deallocated.  Our analysis handles all code
compilable using this version of GCC except for a few cases: we do not handle
accesses that compile to GCC {\tt BITFIELD\_REFERENCE} IR types because these
are undocumented, we do not handle inlined assembly instructions, and we do not
handle C++ exceptions.  Note that these limitations of our research prototype
are not fundamental and the limited cases are uncommon.

We built our runtime system from scratch.  The LWT is a fixed size 2GB array of
64-bit words.  Each entry packs a program point and a thread identifier into
the 64-bit word.  By default, program points are single instruction addresses,
but we also support limited (2-address) context-sensitivity by packing the
current instruction's address and the nearest return address into a 64-bit LWT
entry. When an access to a memory location occurs, we index into the LWT with
the lower bits of the location's address.  Our prototype implementation uses a
lossy resolution policy for hash collisions.  We use this policy in our
prototype because it is fast and bounds memory overheads at 2GB.  Collisions
are rare (see Section~\ref{sec:eval:performance:collisions}), so this
simplification is unlikely to be a problem.

\subsection{Release}
We released our implementation, free and open-sourced, on the
web~\cite{ctrapsrelease} and published it to the GCC plugin repository.
Since its release, we have seen contributions to our codebase from members of
the GCC community and downloads from interested developers.

@@

The results in Table~\ref{tab:bugs} show that for both atomicity and ordering
violation bugs, last writer slices are effective.  In all cases, we found that
the last writer slice revealed the provenance of at least some data involved in
the failure, so last writer slicing was effective.  We found that none of the
failures we looked at were trivially debuggable.  This finding is consistent
with the difficulty of debugging concurrency bugs -- the core dump does not
reveal the interaction between threads that led to a failure, only the result
of the interaction.

Last writer slices were useful for helping understand how to fix the bug in
all but one case and useful at least for understanding the bug in the
remaining case, {\tt apache2}.  The main reason last writer slices are helpful
is that they reveal the connection between code running in different threads
that interacts to cause a failure.  Without last writer slices, programmers
have no easy way of connecting these otherwise disparate parts of a program.
The case studies in Section~\ref{sec:eval:debugging:cases} show in more detail
how last writer slices are useful for debugging and illustrate their limitations.

%For the remaining case, {\tt apache2}, last writer slices would
%probably help with debugging, but it is less obvious because this bug is a
%complex, multi-variable bug.  We provide more detail on {\tt apache2} in
%Section~\ref{sec:eval:debugging:cases}.
Our evaluation shows that bad value origin tracking~\cite{badapples} is
helpful in one case where last writer slices is useful and provides limited help
in one other case.  For a use-after-delete bug in {\tt pbzip2}, bad value
origin tracking helps.  For this bug, when one thread deletes a variable, it
makes that variable unusable and bad value origin tracking would track that
value.  When the failure occurs, the programmer would then know why the value
was deleted.

@@

We used last writer slices to debug an atomicity violation in MySQL-4.0.12 that
is a critical security vulnerability.  Figure~\ref{fig:mysqllws} shows the
buggy code.  Thread 1 rotates the database's log file.  Log rotation should be
atomic, but the lack of synchronization allows Thread 2 to run its insert
during log rotation.  In that case, the insert query is not logged, manifesting
the failure.  We added an {\tt abort} statement to the non-logging case because
this behavior is a security vulnerability that the bug report describes as
``critical''.  Adding the {\tt abort} is reasonable, as the bug report
describes the {\tt else} block as an error case.  Running in a debugger, a
programmer might equivalently use a breakpoint, rather than adding an
{\tt abort}.

@@

\paragraph{apache2}
We used last writer slices to debug an atomicity violation in Apache-2.2.9 and
we found that last writer slices help understand the bug, but not fully how to fix
it.  The bug involves a group of worker threads that handle web requests and a
network listener thread that passes requests to the workers.  The server tracks
the number of idle workers.  When a worker becomes idle, it increments the
number of idlers and when the listener passes work to a worker, it decrements
that number.  The counter's increments and decrements are synchronized using
condition variables.  If there are no idlers, the listener waits for a signal.
When a worker becomes idle when no others are, it delivers a signal to the
listener.

The problem is that the listener does not check the condition governed by the
condition variable after being signaled on that condition variable.  As a
result, the listener can incorrectly decrement the idler count twice,
underflowing the unsigned value, which then causes an assertion to fail.  When
the failure occurs, the core dump shows the idler count underflowed, revealing
a primary symptom of the bug, but not providing much help.  The last
writer slice for the idler count reports that the listener thread last
wrote the idler count at its decrement operation.  That information shows the programmer
{\em why} the underflow occurred, which is helpful for understanding the
failure.  However, we conclude that last writer slices are only somewhat 
useful in this case.  Solving this complex bug requires understanding not just
the connection between the underflow and the listener's decrement, but also the
synchronization protocol that coordinates that decrement and other accesses to
the variable.

@@

These high-level results support our claim that our overheads 
are low enough for deployed systems for many applications, especially
for servers.  We now further analyze our observed overheads, illuminating
differences between the lower-overhead server programs and the higher-overhead
PARSEC programs.

@@

\subsubsection{Overheads Due to Conservative Escape Analysis}
\label{sec:eval:conservative}
{\tt swaptions} had the highest run time overhead in our tests.  We
investigated the cause and determined it is largely due to shortcomings in our
escape analysis.  We located the inner-loop function {\tt
HJM\_SimPath\_Forward\_Blocking}.  The function uses two local matrices that
are allocated through an external call.  Our escape analysis cannot determine
that the matrices are only used locally in that function, so it conservatively
assumes the matrices escape.  We manually inlined the matrix allocation calls
(and a matching deallocation calls), and escape analysis eliminated their
instrumentation.  The program's overhead dropped from about 420\% overhead for
\ctraps to about 300\% overhead; the change for last writer slicing was from
300\% to about 130\% overhead.  We also looked at an output matrix passed into
the inner-loop function.  The matrix is allocated and deallocated in the
inner-loop's caller, but is never shared.  Eliminating instrumentation on
accesses to this matrix yielded an overhead of 46\% for \ctraps and 
just 30\% for last writer slicing.

These manual improvements to the escape analysis reduced the run time overheads
for these workloads from being unacceptable for any production environment to
being reasonable for use in production systems. Better escape analysis would
find these performance gains automatically.

\subsubsection{PARSEC vs. Servers}
\label{sec:eval:parsecserver}
We discern three reasons that server programs better tolerated the overheads of
last writer slices and \ctraps in our experiments.

@@

\subsection{Context-Sensitivity}
%{\bf TODO: Was this LWS or CTRAPS-NOOP or what?}
We briefly evaluated the cost of context-sensitivity using \ctraps with no-op
trap handlers.  With a bounded 2-address context, the average overhead for
server programs was 18\% and the average overhead for PARSEC was 145\%.  These
data show that a restricted form of context-sensitive analysis is possible in
\ctraps with overheads low enough for production in some cases.

\subsection{Memory Overheads}
\label{sec:eval:performance:collisions}
As we describe in Section~\ref{sec:implementation} our prototype \lwt is a 2GB
hashtable of 8 byte entries.  We use a lossy hash collision policy: all updates
always overwrite existing entries.  This strategy fixes memory overheads at
2GB, but frequent lossy collisions may lead to imprecision.   We instrumented
our runtime to count collisions and found that for PARSEC applications the
geometric mean rate of collisions was very low - 89 per 10,000 memory accesses.
This summary result shows that our prototype implementation is reasonable,
especially for debugging in production, when low memory use is key and minor
imprecision is tolerable.  Note that collisions did not introduce noticable
imprecision in our debugging experiments.  When precision is more important than memory overhead, ({\em
e.g.}, for sound, offline analysis), a lossless hashtable is a better option.

@@

\subsubsection{Performance with \ctraps Applications}

@@

\section{Conclusions}
In this work we proposed last writer slicing, the first technique to collect data provenance information
that is not limited to certain values, is targeted to concurrent programs, and
has overheads low enough for production use, especially in server programs.

We developed system support that collects last writer slices with overheads
well within an acceptable range for always-on use in production.  We showed
that during debugging, the provenance information provided by last writer
slices reveals crucial connections between different threads' otherwise
disparate regions of code.  Understanding how code in different threads
interacts is essential to concurrency debugging and this work aids that
understanding.
